\sysfull

### 2018-04-17 paper/paging.tex
fixed 5.4
partial commit on the paging section

\sys exposes non-volatile memory for \emph{protected} variables to
the application as a virtual address space partitioned into pages.
\emph{Paged virtual memory} is a key enabler of task coalescing since it keeps the non-volatile memory consistent despite dynamic WAR dependencies incurred by coalesced code regions. Thanks to memory virtualization, \sys have the freedom to move application's data between volatile and non-volatile memory transparently.
%
%,. Virtualization simplifies
%coalescing, because the system gets the freedom to move application's data
%between volatile and non-volatile memory transparently.
%
To this end, \sys isolates FRAM from the tasks of the application so that they interface only SRAM---which is cheaper to read/write as we indicated in Section~\ref{sec:background}.
%
%manages memory such that application tasks operate primarily from
%efficient volatile SRAM rather than non-volatile FRAM, which is more costly to access than SRAM,
%as we showed in Figure~\ref{fig:framEnergy}. 
\sys runtime implements memory
virtualization by moving pages of the address space from the large,
non-volatile memory to a small, volatile memory on demand and writing all dirty
pages back to non-volatile memory atomically. Since memory moves at page granularity, \sys
exploits hardware\hyp{}accelerated block copies using DMA.

%\subsection{Paged Non-Volatile Virtual Memory}

@@

\textbf{System Overview.} The overall picture of the paging operations is illustrated in Figure~\ref{figure:coala_page_tags}. Paging moves data between non-volatile and volatile memory. A page has a home location in non-volatile memory which we refer to as the page's \texttt{\underline{store}} (underlined locations are non-volatile).
As a task accesses pages, they are swapped from their home location
into the fully\hyp{}associative \texttt{working} page buffer in volatile memory. The \texttt{working} buffer is allocated to take up the volatile memory that remains after space for stack is allocated.
%
\sys uses a two-phase commit to write dirty pages from the \texttt{working} buffer back to their home location in \texttt{\underline{store}} atomically, ensuring data consistency even with power failures.
%
When any dirty page is swapped out from the \texttt{working} buffer due to eviction or task transition from last coalesced task, the first phase writes the page to a direct\hyp{}mapped non-volatile \texttt{\underline{shadow}} buffer, and the second phase writes it from \texttt{\underline{shadow}} to \texttt{\underline{store}}.
%
Unlike prior work~\cite{chain,alpaca} that moves scattered variables one at a time between volatile and non-volatile memory, \sys's page-based design moves contiguous blocks of data.
%
\sys takes full advantage of hardware\hyp{}accelerated block copies using Direct Memory Access (DMA) hardware in common microcontrollers --- e.g. 5$\times$ fewer cycles on MSP430FRxxxx.

@@

A \sys task must access protected non-volatile variables through a restricted memory interface, as will be described in Section~\ref{sec:coala_api}. The interface includes \texttt{RP(var)}, to read the value of variable {\tt var}, and \texttt{WP(var)} to assign a value to variable {\tt var}. The implementation of {\tt RP} is shown in Algorithm~\ref{algo:rwar}, and {\tt WP}'s implementation is similar except for marking the accessed page as dirty.
%
{\tt RP} and {\tt WP} operations translate a variable's physical address in non-volatile memory into a \emph{virtual address} in the page buffer in volatile memory. In \sys, a virtual address is composed of a \emph{page tag} that identifies the page and a \emph{page offset} that identifies a byte.

@@

If the \texttt{working} buffer is full, a page fault on a memory access requires \sys to swap out one of the pages in the volatile buffer, the \emph{victim page}, preserving updates made to that page, and to copy the accessed page of protected data into the page buffer. Algorithm~\ref{algo:pagefault} shows how \sys handles a page fault. A victim page in the \texttt{working} buffer is selected for eviction according to \emph{first-in-first-out} replacement policy (Lines 2 and 12). If the page being swapped out is not dirty and the page being accessed had not been dirtied and swapped out to the \texttt{shadow} buffer since last boot, then handling a page fault consists of copying the accessed page from \texttt{\underline{store}} into the \texttt{working} buffer (Line 11). Otherwise, the page fault handler must take the following extra steps.
%
If a task modified any byte in the victim page (i.e., using \texttt{WP}), then the page is dirty. Before swapping in the accessed page, \sys must first preserve the updated values in the dirty page so that they can be committed back to their locations in memory when the dynamic task ends (but not earlier). To preserve dirty values, \sys copies the victim page to the non-volatile \texttt{\underline{shadow}} buffer (Line 5) and records the page's tag in a non-volatile list (Lines 6-7) to support iteration over shadow pages having traversing all slots in the buffer, necessary in the second phase of the commit.
%
%Dirty pages remain in {\tt \underline{shadow}} buffer until the
%dynamic task completes, at which point \sys commits the modified
%pages from {\tt \underline{shadow}} back to \texttt{\underline{store}}.
%
If the accessed page was previously modified and swapped out since the last boot (Line 8), the most recent version of the page is in {\tt
\underline{shadow}}, not \texttt{\underline{store}} (Line 9).

@@

One design choice \sys made was to insert dynamic check for paging on every
possible problematic memory operation using {\tt RP} and {\tt WP} macros.
Since dynamic checks incur overhead on every read and write, it has a higher
overhead for accessing memory than a static approach,
e.g., Alpaca, where a system statically analyze the memory operation and try to insert only optimal
number of memory protection code (usually at the beginning of each task). 
%
However, \sys's dynamic paging has several benefits over static approach.
First, static approach has limitations, prohibiting certain types of programming pattern. For example,
in the presence of arbitrary pointer operation, function calls using function pointer, or an interrupt
within a task, the system cannot statically analyze the memory behavior. Thus, static
approaches either prohibits the programmer from using certain programming pattern~\cite{alpaca}, 
or will have to conservatively insert memory protection code that can be unnecessary.
Second, static approach cannot handle task coalescing, because with coalescing the scope to be
searched statically is not determined at compile time. 
By choosing to check every memory access dynamically, \sys supports programming pattern that
Alpaca does not support, such as arbitrary pointer operation, while also being able to
support task coalescing, which yields extra performance benefits.
%
%Moreover, if the programmer wants to optimize for high performance, it is possible for the
%programmer to place {\tt RP} and {\tt WP} only minimally by carefully examining the code.
%We argue that the

>>>>

\sys's {\em virtual memory manager} exposes non-volatile memory for
\emph{protected} variables to the application as a virtual address space
partitioned into pages.  
%
A program's tasks read and write these protected variables, using them like any
other data in the program. 
%
The virtual memory manager automatically finds a variable in its page and  
allows a task to directly manipulate its value.
%
When a task completes, the virtual memory manager ensures that updated pages 
are committed to main memory and visible to subesequently executed tasks.
%
If a power failure interrupts a task, the virtual memory manager ensures that
the task sees consistent memory state when it restarts its execution. 
%
The virtual memory manager ensures consistency even if a task or a sequence of tasks include WAR dependences that could compromise consistency~\cite{ratchet,dino}.
%
%
%
%,. Virtualization simplifies
%coalescing, because the system gets the freedom to move application's data
%between volatile and non-volatile memory transparently.
%

\sys's virtual memory manager \sys abstracts non-volatile storage (e.g.,
variables in FRAM) from tasks, which simply manipulate protected variables.
%
Protected variables remain consistent and are durable across power failures.
%
While ensuring variables remain durable, \sys avoids accessing non-volatile
memory when possible, by copying pages of data into a {\em working memory} in the device's SRAM, which
has a lower latency and energy cost to access than the device's FRAM (as we
showed in Section~\ref{sec:background}).
%
%
On a commit, \sys atomically copies all updated (i.e., ``dirty'') pages from their
space in the working memory back into their location in the non-volatile
memory.
%
SRAM has a limited size and there is an extended working memory containing {\em
shadow pages} in the larger FRAM.  A shadow page stores the data from an
updated pages that need to be committed back to non-volatile memory, but that
exceeded the size of the volatile working memory. 
%
%manages memory such that application tasks operate primarily from
%efficient volatile SRAM rather than non-volatile FRAM, which is more costly to access than SRAM,
%as we showed in Figure~\ref{fig:framEnergy}. 
\sys moves pages of data efficiently using hardware accelerated Direct Memory
Access (DMA) support, which copies a block of memory from one place to another
as a single operation.

@@

\textbf{System Overview.} Figure~\ref{figure:coala_page_tags} provides an
overall view of \sys's virtual memory management system.  A page has a home
location in non-volatile memory which we refer to as the page's
\texttt{\underline{store}} (underlined locations are non-volatile).  When a
task accesses a page, the memory manager swaps the page from its home location
into the fully\hyp{}associative \texttt{working} buffer that is stored in
volatile memory.  The \texttt{working} buffer is allocated to take up the
volatile memory that remains after space for stack is allocated.
%
\sys uses a two-phase commit to write dirty pages from the \texttt{working}
buffer back to their home location in \texttt{\underline{store}} atomically,
ensuring data consistency even in the presence of a power failure during commit.
%
The memory manager is forced to swap out a dirty page from the \texttt{working} buffer 
when a task accesses a page that is not in the working buffer, and there is no
free page in the working buffer (i.e., a page fault). On a page fault, \sys
moves the page to a direct\hyp{}mapped non-volatile \texttt{\underline{shadow}}
page buffer. Later, to commit the page, \sys writes it from its
\texttt{\underline{shadow}} location back to its home location in
\texttt{\underline{store}}.
%
%BML: this point is true, but without caches, it is not a performance factor, so I removed it.
%Unlike prior work~\cite{chain,alpaca} that moves scattered variables one at a
%time between volatile and non-volatile memory, \sys's page-based design moves
%contiguous blocks of data.
%
%BML: this seems gratuitous because DMA is very common and we've already introduced it.
%\sys takes full advantage of hardware\hyp{}accelerated block copies using
%Direct Memory Access (DMA) hardware in common microcontrollers --- e.g.
%5$\times$ fewer cycles on MSP430FRxxxx.

@@

A \sys task must access protected non-volatile variables through \sys's
restricted memory interface, which Section~\ref{sec:coala_api} describes in
detail.  The interface includes \texttt{RP(var)}, to read the value of variable
{\tt var}, and \texttt{WP(var)} to assign a value to variable {\tt var}. The
implementation of {\tt RP} is shown in Algorithm~\ref{algo:rwar}, and {\tt
WP}'s implementation is similar except that {\tt WP} marks the accessed page as
dirty.
%
{\tt RP} and {\tt WP} operations translate a variable's physical address in
non-volatile memory into a \emph{virtual address} in the page buffer in
volatile memory. In \sys, a virtual address is composed of a \emph{page tag}
that identifies the page and a \emph{page offset} that identifies a byte.

@@

If the \texttt{working} buffer is full, a page fault on a memory access
requires \sys to swap out one of the pages in the volatile buffer (a 
\emph{victim page}) preserving updates made to that page, and to copy the
accessed page of protected data into the working buffer.
Algorithm~\ref{algo:pagefault} shows how \sys handles a page fault. A victim
page in the \texttt{working} buffer is selected for eviction according to
a \emph{first-in-first-out} replacement policy (Lines 2 and 12). If the page
being swapped out is not dirty and the page being accessed had not been dirtied
and swapped out to the \texttt{shadow} buffer since last boot, then handling a
page fault consists of copying the accessed page from
\texttt{\underline{store}} into the \texttt{working} buffer (Line 11).
Otherwise, the page fault handler must take the following extra steps.
%
If a task modified any byte in the victim page (i.e., using \texttt{WP}), then
the page is dirty. Before swapping in the accessed page, \sys must first
preserve the updated values in the dirty page so that they can be committed
back to their locations in memory when the dynamic task ends (but not earlier).
To preserve dirty values, \sys copies the victim page to the non-volatile
\texttt{\underline{shadow}} buffer (Line 5) and records the page's tag in a
non-volatile list (Lines 6-7) to support iteration over shadow pages having
traversing all slots in the buffer, necessary in the second phase of the
commit.
%
%Dirty pages remain in {\tt \underline{shadow}} buffer until the
%dynamic task completes, at which point \sys commits the modified
%pages from {\tt \underline{shadow}} back to \texttt{\underline{store}}.
%
If the accessed page was previously modified and swapped out since the last boot (Line 8), the most recent version of the page is in {\tt
\underline{shadow}}, not \texttt{\underline{store}} (Line 9).

@@

One design choice \sys made was to insert dynamic check for paging on every
possible problematic memory operation using {\tt RP} and {\tt WP} macros.
Since dynamic checks incur overhead on every read and write, it has a higher
overhead for accessing memory than a static approach,
e.g., Alpaca, where a system statically analyze the memory operation and try to insert only optimal
number of memory protection code (usually at the beginning of each task).

However, \sys's dynamic paging has several benefits over static approach.
First, static approach has limitations, prohibiting certain types of programming pattern. For example,
in the presence of arbitrary pointer operation, function calls using function pointer, or an interrupt
within a task, the system cannot statically analyze the memory behavior. Thus, static
approaches either prohibits the programmer from using certain programming pattern~\cite{alpaca}, 
or will have to conservatively insert memory protection code that can be unnecessary.
Second, static approach cannot handle task coalescing, because with coalescing the scope to be
searched statically is not determined at compile time. 
By choosing to check every memory access dynamically, \sys supports programming pattern that
Alpaca does not support, such as arbitrary pointer operation, while also being able to
support task coalescing, which yields extra performance benefits.
%
%Moreover, if the programmer wants to optimize for high performance, it is possible for the
%programmer to place {\tt RP} and {\tt WP} only minimally by carefully examining the code.
%We argue that the
